---
title: "PCA in R for Decathlone Dataset"
author: "Anjum Ismail"
date: "28/03/2021"
output: html_document
---

# Download all the libraries rerquied

```{r include=FALSE}
if(!require(skimr)) install.packages("skimr") #skim
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(FactoMineR)) install.packages("FactoMineR") #prcomp
if(!require(factoextra)) install.packages("factoextra") # get_eig() fviz_eig()
if(!require(mlbench)) install.packages("mlbench") # contains various datasets
if(!require(caret)) install.packages("caret") #createDataPartition
if(!require(RANN)) install.packages("RANN")
if(!require(mice)) install.packages("mice")
if(!require(NbClust)) install.packages("NbClust")

library(skimr)
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(mlbench)
library(caret)
library(RANN)
library(mice)
library(NbClust)
```


# Principal Component Analysis

The dataset `decathlon` contains athletes' performance during two sporting meetings. The first ten columns corresponds to the performance of the athletes for the 10 events of the decathlon. The columns 11 and 12 correspond respectively to the rank and the points obtained. The last column is a categorical variable corresponding to the sporting event (2004 Olympic Game or 2004 Decastar).


```{r}
# load the dataset
data ("decathlon")

# Create a copy of dataset
decathlon_copy <- decathlon

# Inspect the dataset
head(decathlon)

```


```{r}
# Inspect the broader overview of dataset
skim (decathlon)
```
The above result gives a clear view that the dataset does not have any missing values


```{r}
# To Stimulate the real world scenario lets divide the dataset into training and test.
# The training dataset will be 90% and the remaining 10% is test data set
# Feature Rank and point is ignored 
# First 10 features will be the input and Competition will be the target feature

train_data <- createDataPartition(decathlon_copy$Competition, p=0.9, list=FALSE)
decathlon_train <- decathlon_copy[train_data, ]
decathlon_test <- decathlon_copy[-train_data, ]

# verify the dimensions of test and train data set created
dim_train<- as.vector(dim (decathlon_train))
dim_test<- as.vector(dim (decathlon_test))

paste0 ("Training dataset has ", dim_train[1], " records")
paste0 ("Test dataset has ", dim_test[1], " records")

```
```{r}
# Applying PCA for the first 10 columns and also performing scaling
# From the proportion of variance we can see that PC1 is having having the highest variance with respect to data
decathlon_pca <- prcomp(decathlon_train[, 1:10], center = TRUE, scale. = TRUE)
summary(decathlon_pca)

#Displaying the eigen vectors for every PCA and feature
decathlon_pca$rotation

```
```{r}
# Extract the Eigen values and identify the number of PC's to be considered for 80% of over all variability

eigen_values <- get_eig(decathlon_pca)
eigen_values

# From the below result we can confirm that PC1 to PC% should be considered for 80% variability 
```

```{r}
# Scree plot the vairance with respect to PC

fviz_eig(decathlon_pca)

# From the below plot it is very clear that there is very little variability after PC5
```



```{r}
# Inspect the variance of the data with respect to PC

result <- get_pca_var(decathlon_pca)
result

```

```{r}
# Let identify the correlation between the each variable and the PCi

res_cor<- result$cor
res_cor

```

```{r}
# Identify the proportion of variance between each variance and its PCi

res_cos2 <- result$cos2
res_cos2
```

```{r}
# Let us try to visualize the variance for the first 2 PCi
result_plot <- cbind (res_cos2[,1:2], variability_explained = rowSums(res_cos2[,1:2]))
result_plot

fviz_pca_var(decathlon_pca, col.var = "cos2",gradient.cols = "lancet")

# From the below visualization we observe that variable "400m" has the highest variance wrt the PCi

```

```{r}

# Plot all the observation against the first two PCs
# Colour individuals by the competition type


fviz_pca_biplot(decathlon_pca, label="var", habillage = decathlon_train$Competition, addEllipses = TRUE)

```

```{r}

# Creating the updated training dataset with PC1 to PC5 because the overall 80% vraibility reflects within PC1 to PC5

decathlon_train_pca_transformed <- decathlon_pca$x[ ,1:5] %>%
  as.data.frame() %>%
  mutate (Competition = decathlon_train$Competition)

decathlon_train_pca_transformed


```


```{r}

# Predict the test result for the PCA mode created 

decathlon_test_pca <- predict(decathlon_pca, decathlon_test[, 1:10])

decathlon_test_pca_transformed <- decathlon_test_pca[,1:5] %>%
  as.data.frame() %>%
  mutate(Competition = decathlon_test$Competition)

decathlon_test_pca_transformed

```

